Determining the existance and direction of causality in the face of measurement error

### 2016-05-08 docs/manuscript.Rmd
updated

fig_nums(
	"d_relationship_figure", 
	caption="Plots depicting the function $d = cor(x, x_O) - cor(x,y)cor(y, y_O)$.",
	display=FALSE
)

@@

### The influence of measurement error on CIT

@@

### Inferring direction of causality using MR

@@

thus we obtain an estimate of the effect of $x_O$ on $y$ that is scaled by the measurement error in $x$ and $y$, but is unrelated to measurement imprecision. Measurement error and imprecision are likely to reduce the power of MR inferring a causal association (increased false negative rate, but importantly the evidence for causal association, based on rejecting the null hypothesis that $\beta_{MR} = 0$, is not inflated when the null hypothesis is true.

However, if we do not know whether the SNP $g$ has a direct influence on $x$ or $y$ then further efforts are required. If $x$ causes $y$ and $g$ is a valid instrument for $x$ then we expect that the Steiger test will obtain the correct result. This is because the value $d = \rho_{g, x_O} - \rho_{g, y_O}$ is greater than 0 when the causal association between $x$ and $y$ is $\rho_{x, y} < 1$. But it can be shown that in the presence of measurement error, $d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}$ (Appendix 2), thus it is important to note that under certain conditions when measurement error is present the Steiger test could make erroneous inference about causal direction because the value of $d$ is not restricted to being greater than 0. However, using the MR approach there are two potential advantages over the CIT. First, we can conduct a formal test for the direction of causality. Second the correct direction of causality can be inferred when measurement error in the exposure variable is present, as long as it is lower than the product of the measurement error in the outcome and the causal correlation between the exposure and the outcome. `r cfig("d_relationship_figure")` shows that in most cases, especially when the causal effect between $x$ and $y$ is not very large, the condition of $d > 0$ is satisfied.

>>>>

fig_nums(
	"d_relationship_figure", 
	caption="Plots depicting the function $d = cor(x, x_O) - cor(x,y)cor(y, y_O)$. $d$ is plotted on the y axis, with measurement precision of $x$ ($cor(x,x_O)$) on the x-axis. Higher measurement precision of $y$ ($cor(y,y_O)$) is depicted by lighter coloured lines, and columns of boxes are for different true values of $cor(x,y)$. The graph shows that for the majority of the parameter space of the function, $d$ is positive, especially where causal relationships are relatively weak.",
	display=FALSE
)

fig_nums(
	"causality_exists_tpr",
	caption="Outcomes were simulated to be causally influenced by exposures with varying degrees of measurement imprecision applied to the exposure variable (x axis). True positive rates (y axis) for MR and CIT were compared with for varying levels of measurement imprecision in the outcome variable (rows of boxes) and sample sizes (columns of boxes)."
)

@@

### The influence of measurement error on mediation

@@

### Inferring the existence of causality

@@

Thus we obtain an estimate of the effect of $x_O$ on $y$ that is scaled by the measurement error in $x$ and $y$, but is unrelated to measurement imprecision. Measurement error and imprecision are likely to reduce the power of MR inferring a causal association (increased false negative rate), but importantly the evidence for causal association, based on rejecting the null hypothesis that $\beta_{MR} = 0$, is not inflated when the null hypothesis is true.

We performed simulations to compare the performance of MR against CIT in detecting a causal association between simulated variables under different levels of imprecision simulated in the exposure. `r cfig("causality_exists_tpr")` shows the true positive rates between the CIT and MR for detecting a causal association. We observe that the CIT has lower power in all cases, with performance declining as measurement imprecision increases in the exposure. The performance of MR in detecting an association is unrelated to measurement error in. Measurement error in the outcome does not induce a substantive difference in performance between CIT and MR.

### Inferring the direction of causality

If we do not know whether the SNP $g$ has a direct influence on $x$ or $y$ then further efforts are required. If $x$ causes $y$ and $g$ is a valid instrument for $x$ then we expect that the Steiger test will obtain the correct result. This is because the value $d = \rho_{g, x_O} - \rho_{g, y_O}$ is greater than 0 when the causal association between $x$ and $y$ is $\rho_{x, y} < 1$. But it can be shown that in the presence of measurement error, $d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}$ (Appendix 2), thus it is important to note that under certain conditions when measurement error is present the Steiger test could make erroneous inference about causal direction because the value of $d$ is not restricted to being greater than 0. However, using the MR approach there are two potential advantages over the CIT. First, we can conduct a formal test for the direction of causality. Second the correct direction of causality can be inferred when measurement error in the exposure variable is present, as long as it is lower than the product of the measurement error in the outcome and the causal correlation between the exposure and the outcome. `r cfig("d_relationship_figure")` shows that in most cases, especially when the causal effect between $x$ and $y$ is not very large, the condition of $d > 0$ is satisfied.
